---
title: "Principle Component Analysis"
description: |
  A way of understanding variable relationships
author:
  - name: Germán Silva
    url: {}
date: 2022-03-12
output:
  distill::distill_article:
    self_contained: false
---

# Overview

**Summary**:

This report was done as part of UCSB's ESM 244 *Advanced Data Analysis* and looks at miscellaneous environmental variables round the world at the country level from Zander Venter on Kaggle. I explore the relationships between some of these variables by using principle component analysis (PCA). Variables were selected the minimize the number of rows removed by NA deletion. Final selected variables include accessibility to cities, cropland cover, tree canopy cover, annual mean rainfall and temperature, and cloudiness. The relationships between these variables are interpreted from a biplot and screeplot created from the PCA results. 

**Data Source**

Data compiled by Zander Venter (2018). "Environmental variables for world countries". Accessed from: https://www.kaggle.com/zanderventer/environmental-variables-for-world-countries

# Principle Component Analysis

## Set up code:

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

## attach libraries

library(tidyverse)
library(here)
library(ggfortify)
```

## Read in the data:

```{r}
# Read in the data
world_env <- read_csv(here("data", "world_env_vars.csv")) %>% # open csv
  select(Country,
         accessibility_to_cities, 
         cropland_cover, 
         tree_canopy_cover, 
         rain_mean_annual, 
         temp_mean_annual, 
         cloudiness) %>% # select the variables we want to explore
  drop_na() # drops columns with NAs as PCA only works with numeric values
```

## Perform the PCA:

PCA is easily performed in `R` with the `prcomp()` function. Scaling is performed to avoid any potential issues with comparisons between variables with different units. As PCA can only be done on numeric variables, the country names were removed for this step. 

```{r}
# Performing the PCA
world_env_pca <- world_env %>% # pipe in the data
  select(!Country) %>% # remove the country
  prcomp(scale. = TRUE) # rescaling of data is performed

```

## Create biplot and screeplot:

A biplot is one way to interpret the results from a PCA. The arrows indicate the strength and correlation of the relationships of the variables to the principle components (PCs). The direction of the arrows relative to each other give a sense of their correlation or lack thereof in the axes determined by the different PCs.

```{r, fig.align='center', fig.cap="Biplot showing relationships between variables in PC1 and PC2. A few relationships between the variables stick out: 1) cloudiness, tree canopy cover, and mean annual rainfall seem to be positively correlated, 2) mean annual temperature and cropland are strongly negatively correlated with the angle between them being near 180^o^, 3) cropland cover and mean annual temp seem to have no to a weak correlation with cloudiness, tree canopy cover, and mean annual rainfall."}

#Creating the biplot via autoplot()

autoplot(world_env_pca, # pca data
         data = world_env, # original data
         loadings = TRUE, # show loadings
         loadings.label = TRUE, # label loadings
         loadings.colour = "black", # loading color
         loadings.label.colour = "black", # loading label color
         loadings.label.vjust = -.5) + # vertical justification
  ggtitle("Principle Component 1 and 2 Biplot (SILVA)") + # title
  labs(x = "Principle Component 1 (43.47%)", 
       y= "Principle Component 2 (23.42%)")+ # axis labels 
   theme(plot.title = element_text(color = "#5b4f41", size = 16),
            plot.background = element_rect("white"),
            panel.background = element_rect("#faf7f2"),
            panel.grid = element_line(linetype= "longdash",
                                      color = "#f0ece1"),
            axis.text = element_text(color = "#5b4f41"),
            axis.title = element_text(color = "#5b4f41"),
            strip.background = element_rect("white"),
            axis.line = element_line(color = "#5b4f41")) # change themes
```

Screeplots allow us to see how much of the variance in the data is explained by each of the PCs. Typically when ~80% of the variance is explained by the PCs is good. 

```{r,fig.align='center', fig.cap="Screeplot showing the amount of variance explained by each principle component."}
# Creating screeplot
sd_vec <- world_env_pca$sdev # standard deviation
var_vec <- sd_vec^2 # variance

pc_names <- colnames(world_env_pca$rotation) # names of PCs

pct_expl_df <- data.frame(v = var_vec, # new data frame for screeplot; variance
                          pct_v = var_vec / sum(var_vec), # percent of variance
                          pc = fct_inorder(pc_names)) %>%  # orders rows
  mutate(pct_label = paste0(round(pct_v * 100, 1), "%")) # adds character %

ggplot(pct_expl_df, aes(x = pc, y = v))+ # graphs via PC and variance
  geom_col()+
  geom_text(aes(label = pct_label), vjust = 0, nudge_y = 0.008)+
  labs(x = "Principle Components",
       y = "Variances") +
  ggtitle("PCA Screeplot")+
  theme(plot.title = element_text(hjust = .5))
```

# Takeaways:

- Cloudiness, tree canopy cover, and mean annual rainfall have a positively correlated relationship with each other. 
  - This makes conceptual sense as countries with more rain would have more cloudy days and likely be able to support more trees.
- Mean annual temperature and cropland have a strong negatively correlated relationship with the angle between them being near 180^o^.
  - We might expect hotter countries (like Saudi Arabia) to have less arable land due to the extreme heat due to other environmental factors. 
- Cropland cover and mean annual temp seem to be not to weakly correlated with cloudiness, tree canopy cover, and mean annual rainfall. 
- The first two principle components account for 66.9% of the total variance in the data with over 80% being explained by the first three principle components. 
