---
title: "Spatial Data Visualization and Point Pattern Analysis"
description: |
  Spatial data and knowing how to visualize and analyze it is an important skill in geography
author:
  - name: Germán Silva
    url: {}
date: 2022-03-12
output:
  distill::distill_article:
    self_contained: false
---

# Overview

**Summary**:

This entry highlight spatial data visualization done and point pattern analysis done as part of UCSB's ESM 244. This code provides two examples of spatial data visualization for oil spill data in California and point pattern analysis to assess whether the spills are clustered or not. The exploratory data visualization is in an interactive format, while the choropleth map is a static map in a finalized format. Both highlight how the same data can be shown in different ways to visualize different aspects of the spatial data. For example, the full data set is likely useful for interactive exploration, but would make for a messy map if we were trying to convey counts in an easy to understand format. A choropleth makes the data easy to interpret, but obscures where exactly the events occurred. 



**Data Citation**:
CA Department of Fish and Wildlife, Office of Spill Prevention and Response, "Oil spill Incident Tracking [ds394]," https://gis.data.ca.gov/datasets/7464e3d6f4924b50ad06e5a553d71086_0/explore?showTable=true

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, results = FALSE)

# attach libraries
library(spatstat)
library(tmap)
library(maptools)
library(sf)
library(tidyverse)
library(here)
```

# Exploratory Map

## Read in the data

We read in the data for the oil spills and CA counties as simple features (or "sf") using `read_sf()`. 

```{r}
# spill data
spills <- read_sf(here('data', 'Oil_Spill_Incident_Tracking_[ds394]',
                       "Oil_Spill_Incident_Tracking_[ds394].shp")) %>% 
  janitor::clean_names()

# county shapefile
ca_counties <- read_sf(here('data', 'ca_counties', 
                            'CA_Counties_TIGER2016.shp')) %>% 
  janitor::clean_names()
```

## Create interactive exploratory map

We need to first explore the data to get a sense of where things lie in geographic space and how we may want to analyse the data. Here we accomplish this through an interactive map.

```{r, fig.align='center'}
# interactive mode
tmap_mode('view')

# creation of exploratory map with both layers
tm_shape(ca_counties)+
  tm_fill(col = 'black', alpha = 0.3)+
  tm_borders(col = 'black', lwd = 0.5)+
  tm_shape(spills) +
  tm_dots()
```

# Creation of visualization for 2008 Inland Spills by County

## Subset and join data

`sf` objects can be used much like a data frame in `R` and this allows us to use typical data wrangling technqiues on them. Here we subset the oil spill point layer for events that occurred inland in 2007 and spatially join it with the CA counties layer.

```{r}
# subset of data
spills_inland <- spills %>% 
  filter(inlandmari == "Inland") %>% 
  mutate(dateofinci = lubridate::ymd(dateofinci),
         year = lubridate::year(dateofinci)) %>% 
  filter(year == 2008)

# spatial joining
ca_spills <- ca_counties %>% 
  st_join(spills_inland)
```

## Obtain the number of 2008 spills by county

Once the two layers are joined, we can use `summarize()` to obtain a count of the number of oil spills by county.

```{r}
ca_spill_count <- ca_spills %>% 
  group_by(name) %>% 
  summarize(n_records = sum(!is.na(oesnumber)))
```

## Create ggplot for visualization

Once we have the number of oil spills by county, we can create a choropleth that will visualize that data in an easy to interpret static map.

```{r, fig.align='center', fig.cap= "Map of California Counties colored according to the number of oil spills that occured in 2008. Lighter colors indicate fewer icidents than darker colors."}
ggplot(ca_spill_count, aes(fill = n_records))+
  geom_sf() +
  scale_fill_viridis_c(option = "A",direction = -1)+
  theme(plot.title = element_text(color = "#5b4f41", size = 16),
            plot.background = element_rect("white"),
            panel.background = element_rect("#faf7f2"),
            panel.grid = element_line(linetype= "longdash", color = "#f0ece1"),
            axis.text = element_text(color = "#5b4f41"),
            axis.title = element_text(color = "#5b4f41"),
            strip.background = element_rect("white"),
            axis.line = element_line(color = "#5b4f41"))+
  labs(x = "Latitude",
       y = "Longitude",
       fill = "# of Spills")+
  ggtitle("2008 California Oil Spills by County")
```

# Point Pattern Analysis (G-function)

We can also perform point pattern analysis on the data to see understand more about the spatial distribution of the data. This can be done using a G-function to assess nearest neighbors between points. 

```{r, fig.align='center', fig.cap="A G-function plot used assess whether events are spatially random or not. As the G(r) values (% of Points pairs with a neighbor within the distance) are higher for the observed values than the model, we can conclude that CA oil spills are a clustered phenomena."}
spills_sp <- as(spills, "Spatial")
spills_ppp <- as(spills_sp, "ppp")

ca_sp <- as(ca_counties, "Spatial")
ca_win <- as(ca_sp, "owin")

spills_full <- ppp(spills_ppp$x, spills_ppp$y, window = ca_win)

r_vec <- seq(0, 25000, by = 100) # make a sequence of distances for G(r)


gfunction <- envelope(spills_full, fun = Gest, r = r_vec, nsim = 10, nrank = 2)

gfunction_long <- gfunction %>% 
  as.data.frame() %>% 
  pivot_longer(cols = obs:hi, names_to = "model", values_to = "g_val")

ggplot(gfunction_long, aes(x = r, y = g_val, group = model))+
  geom_line(aes(color = model)) +
  scale_color_manual(values = calecopal::cal_palette(name = "sierra1",
                                                     n = 4, 
                                                     type = "discrete"))+
  theme(plot.title = element_text(color = "#5b4f41", size = 16),
            plot.background = element_rect("white"),
            panel.background = element_rect("#faf7f2"),
            panel.grid = element_line(linetype= "longdash", color = "#f0ece1"),
            axis.text = element_text(color = "#5b4f41"),
            axis.title = element_text(color = "#5b4f41"),
            strip.background = element_rect("white"),
            axis.line = element_line(color = "#5b4f41"))+
  labs(x = "Distance",
       y = "% of Point Pairs",
       color = "Model")+
  ggtitle("Nearest Neighbor Point Pattern Analysis by G-Function")

```

